Scalable Clustering of Genotype Information using MapReduce

### 2015-03-11 doc/publications/journal2014/paper.tex
More small changes

Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.  Traditionally both the number of individuals and included genotypes, typically from SNP arrays, were relatively small and libraries in Bioconductor sufficient. However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible. It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 genomes project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}. At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude. This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing.

@@

To demonstrate the capability we cluster variant datasets from the 1000 genomes project to determine population structure using the k-means clustering algorithm available in Mahout. In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.

@@

While the overall processing time is already 40 minutes faster in Mahout than R, the real bottleneck for larger dataset will be the amount of memory available on a single machine. 
As an example, the entire set of 1000 Genome Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}). This is based on the equation:
$\frac{n \times m \times 8\text{bytes}}{1\text{GB}}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a double occupies (i.e. each matrix element), and  1GB is $10^9$bytes.

@@

\subsection*{Clustering quality}
We scored the clusters with the \ARI{}~\cite{Hubert1985}. 
This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$. 
The \ARI{} is then 
{\tiny
\begin{eqnarray*}
ARI=\frac{\sum_{ij}{{n_{ij}\choose 2}} - \left[ \sum_{i}{{a_i\choose2}} \sum_{j}{{b_i\choose2}} \right] / {n\choose2}}{\frac{1}{2} \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] - \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] / {n\choose2}} 
\end{eqnarray*}
}
where $a$ and $b$ are the sums over the rows and columns of the contingency table.

The value is zero for independent clusterings and one for identical clusterings. 
We compared the Mahout k-means clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project.
For the known clusters, individuals from the both phase1 and phase3 datasets were partitioned by their super population code, ASN, EUR, AFR, and AMR, where phase3 included EAS and SAS instead of ASN.

@@

\begin{table}[!t]
\processtable{Datasets used in this study.\label{Tab:01}}
{\begin{tabular}{lrrrrr}\toprule
name& individuals & variants & in-memory  & population\\
& & &size (GB) &groups& \\\midrule
        \NinteenPhaseone{} & 1,092 & 816,115 & 7.13  & 4\\
        \OnePhaseone{} & 1,092 & 3,007,196 & 26.27  & 4\\
        \SevenPhaseone{} & 1,092 & 18,984,880 & 165.85 & 4\\
        \FullPhaseone{} & 1,092 & 38,219,238 & 333.88 & 4\\
	\OnePhasethree{} & 2,504 & 6,468,094 & 131.17 & 5\\
	\ThreePhasethree{} & 2,504 & 19,381,970 & 393.06 & 5\\
	\FullPhasethree{} & 2,504 & 81,271,745 & 1602.00 & 5\\\botrule
\end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 genomes project.
In-memory size is the approximate memory required to store the data as a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
}
\end{table}

>>>>

Grouping individuals based on the genomic profile is a commonly performed tasks to identify population structure~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.  Traditionally both the number of individuals and included genotypes, typically from SNP arrays, were relatively small and libraries in Bioconductor sufficient. However, recent technological advances in whole genome sequencing have made population-scale sequencing feasible. It is hence economical to generate studies with sample sizes currently reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or the cancer genome atlas (TCGA)~\cite{TCGA2013}. At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude. This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing.

@@

To demonstrate the capability we cluster variant datasets from the 1000 Genomes Project to determine population structure using the k-means clustering algorithm available in Mahout. In the first section we benchmark the performance and accuracy of Mahout's implementations against a standard R based implementation on a reduced version of the data and in the last two sections we investigate how Mahout scales to either more samples or more features per samples.

@@

While the overall processing time is already 40 minutes faster in Mahout than R, the real bottleneck for larger dataset will be the amount of memory available on a single machine. 
As an example, the entire set of 1000 Genomes Project phase3 variants could require approximately 1.6TB to be completely loaded into memory as a double-precision matrix (see Table~\ref{Tab:01}). This is based on the equation:
$\frac{n \times m \times 8\text{bytes}}{1\text{GB}}$, where $n \times m$ are the matrix dimensions, 8bytes is the memory a double occupies (i.e. each matrix element), and  1GB is $10^9$bytes.

@@

\subsection*{Clustering quality}
We scored the clusters with the \ARI{}~\cite{Hubert1985}. 
This index compares two different cluster sets, here the prediction $X = \{ X_1, X_2, \ldots , X_r \}$ and the annotated data $Y = \{ Y_1, Y_2, \ldots , Y_s \}$, and assigns a value based on the overlap between $X$ and $Y$ as captured in a contingency table $\left[n_{ij}\right]$ where each entry $n_{ij}$ denotes the number of objects in common between $X_i$ and $Y_j : n_{ij}=|X_i \cap Y_j|$. 
The \ARI{} is then 
{\tiny
\begin{eqnarray*}
ARI=\frac{\sum_{ij}{{n_{ij}\choose 2}} - \left[ \sum_{i}{{a_i\choose2}} \sum_{j}{{b_i\choose2}} \right] / {n\choose2}}{\frac{1}{2} \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] - \left[ \sum_{i}{{a_{i}\choose 2} + \sum_{j}{{b_{j}\choose 2}}} \right] / {n\choose2}} 
\end{eqnarray*}
}
where $a$ and $b$ are the sums over the rows and columns of the contingency table.

The value is zero for independent clusterings and one for identical clusterings. 
We compared the Mahout k-means clusters and the R k-means clusters to the known clusters based on meta data from the 1000 Genomes Project.
For the known clusters, individuals from the both phase1 and phase3 datasets were partitioned by their super population code, ASN, EUR, AFR, and AMR, where phase3 included EAS and SAS instead of ASN.

@@

\begin{table}[!t]
\processtable{Datasets used in this study.\label{Tab:01}}
{\begin{tabular}{lrrrrr}\toprule
name& individuals & variants & in-memory  & population\\
& & &size (GB) &groups& \\\midrule
        \NinteenPhaseone{} & 1,092 & 816,115 & 7.13  & 4\\
        \OnePhaseone{} & 1,092 & 3,007,196 & 26.27  & 4\\
        \SevenPhaseone{} & 1,092 & 18,984,880 & 165.85 & 4\\
        \FullPhaseone{} & 1,092 & 38,219,238 & 333.88 & 4\\
	\OnePhasethree{} & 2,504 & 6,468,094 & 131.17 & 5\\
	\ThreePhasethree{} & 2,504 & 19,381,970 & 393.06 & 5\\
	\FullPhasethree{} & 2,504 & 81,271,745 & 1602.00 & 5\\\botrule
\end{tabular}}{Datasets used in this study; based on portions of the phase1 and phase3 data from the 1000 Genomes Project.
In-memory size is the approximate memory required to store the data as a double-precision matrix of size n,m using the formula: n*m*8bytes/10\textsuperscript{9}bytes.
}
\end{table}
